Evaluating infectious disease forecasts with allocation scoring rules

### 2023-12-20 alloscore_manuscript/alloscore-application-shorter.Rnw
Assorted main text (shorter) edits

\usepackage{amsmath, amsfonts, amssymb}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{amsthm}
\usepackage{bm}
\usepackage{cases}
\usepackage{caption}
\usepackage{xr-hyper}
\usepackage{hyperref}
\usepackage{xurl}
\externaldocument{supplement}

@@

We begin with an informal description of the allocation score and some examples illustrating its key characteristics in section \ref{sec:methods.overview}. In section \ref{sec:methods.detailed} we develop the allocation score more carefully, building on decision theoretic procedures for deriving proper scoring rules. In section \ref{sec:methods.related}, we note that another group of common scores including the quantile score, WIS, and CRPS, can also be derived from decision theoretic foundations \textemdash starting from a different decision making context.

@@

Suppose that a decision maker is tasked with determining how to allocate $K$ available units of a resource across $N$ locations.
If the decision maker is provided with a multivariate forecast $F$ where each marginal forecast distribution $F_i$ predicts resource need in a particular location, one option is to choose the resource allocation that minimizes the expected total unmet need according to the forecast.
We will give a more precise mathematical statement in section \ref{sec:methods.detailed}, but informally, the total expected unmet need according to the forecast is
\begin{align}
\sum_{i=1}^N \mathbb{E}_{F_i}[\text{unmet need in location $i$}], \label{eqn:informal_objective}
\end{align}
where the unmet need in a particular location is the difference between resource need in that location and the number of resources that were allocated there.
This allocation problem has an intuitively appealing solution: allocate so that the probabilities of need exceeding allocation in various locations are as close to each other as possible.
This will lead to the allocations provided by $F$ being quantiles of the marginal distributions $F_i$ for some \emph{single} probability level $\tau$ that is shared in common for all locations.

@@

We can formalize the preceding discussion with the following three-step procedure for developing scoring rules for probabilistic forecasts:
\begin{enumerate}
\item Specify a \emph{loss function} $s(x, y)$ that measures the loss associated with taking action $x$ when outcome $y$ eventually occurs.
\item Given a probabilistic forecast $F$, determine the \emph{Bayes act} $x^F$ that minimizes the expected loss under the distribution $F$.
\item The \emph{scoring rule} for $F$ calculates the score as the loss incurred when the Bayes act was used: $S(F, y) = s(x^F, y)$.
\end{enumerate}
This is a general procedure that may be applied in settings where it is possible to specify a quantitative loss function. Subject to certain technical conditions, scoring rules obtained from this procedure are proper. We refer the reader to the supplement for a more technically precise discussion.

@@

To make this a more easily interpreted measure of forecast performance, we will adjust the raw score by subtracting the minimum loss $l_{\text{oracle}}(y;K)$ achievable by an \emph{oracle} allocator which has precise foreknowledge of the outcomes $y_i$.
When the oracle has sufficient resources to meet the total need, i.e., when $\sum_{i=1}^{N}y_i \leq K$, the oracle's loss is zero and there is no adjustment.  Thus the allocation score coincides with
the raw score in that case.
On the other hand, when the oracle cannot cover all need and incurs a loss of $L \cdot \left(\sum_{i=1}^{N}y_i - K \right) > 0$, we adjust
the raw score by this loss.
The oracle-adjusted score is
\begin{align}
S_A(F, y; K)  &= S_A^{\text{raw}}(F, y; K) - l_{\text{oracle}}(y;K) \\
&= \begin{cases}
S_A^{\text{raw}}(F, y; K), &  \sum_{i=1}^{N}y_i \leq K \\
S_A^{\text{raw}}(F, y; K) - L \cdot \left( \sum_{i=1}^{N}y_i - K\right), &  \sum_{i=1}^{N}y_i > K.
\end{cases}
\end{align}
The oracle adjustment aligns with a common theme in economic decision theory that \emph{opportunity loss} (often known as \emph{regret} or (negative) \emph{relative utility}) is often a more important quantity than absolute loss (see e.g., \cite{DIECIDUE201788}).

@@

In epidemiological forecasting, well-known proper scoring rules such as the log score or variations on the continuous ranked probability score (such as the weighted interval score, or WIS) have been frequently utilized to evaluate probabilistic forecasts.
Often, these scores are used to rank models by accuracy according to that particular score, but without reference to the underlying decision-making process for which that score was designed.
With careful thought and collaboration between modelers and public health officials, we argue that scores that are more aligned with public health decisions could be developed to inform specific problems.
We have demonstrated that forecast evaluation methods that are tied to a specific decision making context can yield model rankings that are substantively different from standard measures of forecast skill.

@@

In practice, there are many potential users of forecasts with many different decision making problems.
Not all can be easily quantified.
Those that can be easily quantified may differ enough that no single score would be appropriate for all users.
Ideally, targeted forecasting tools could be developed through close collaboration between modelers and public health officials.
However, this may only be possible in settings with sufficient staffing on both an analytics and a public health team.
Increasingly, collaborative modeling hubs are being used to generate ``one-size-fits-all'' forecasts for many locations at once.
In these settings, where tailored models are not available, it still could be possible to evaluate contributed models using a set of multiple scores to support public health end users in understanding the value of forecasts as an input to their particular decision making contexts.
% NGR: leaving this out for now as it feels like too much detail?
% This may be tricky to operationalize in the setting of a general forecast hub. It matters how you elicit and represent probabilistic forecasts (quantiles? samples? cdfs?).

@@

There are several important limitations to the current work.
The allocation score we developed here does not directly account for important considerations such as fairness or equity of allocations.
Nor does the proposed framework attempt to capture the broader context of decision making. For example, in practice it may be possible to increase the resource constraint $K$ by shifting funding from other disease mitigation measures.
We also note that in some settings, a ``successful'' epidemiological forecast may lead to policy decisions that change the distribution of the predicted outcome $Y$. Our framework cannot be used to evaluate a forecast at a horizon where the outcome may have been impacted by such a decision.

@@

In conclusion, we argue that the way modelers and policymakers view and evaluate forecasts should change depending on the specific decision-making context.
Using standard forecast evaluation metrics can mask the utility of certain forecasts, or lead to forecasts being used to inform a particular decision that are not well matched to that decision making context.
New collaborative work between public health officials and modeling teams is needed to assess the value and relevance of the initial findings presented here, including real-time pilot studies or simulation exercises that could be used to inform further development of new or alternative scoring metrics.
We see this work as an initial overture for what we hope will grow to be a large, collaborative body of work more closely coupling applied epidemiological forecasting with public health decision making.

>>>>

\usepackage{amsmath, amsfonts, amssymb}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{amsthm}
\usepackage{bm}
\usepackage{cases}
\usepackage{caption}
\usepackage{xr-hyper}
\usepackage{hyperref}
\usepackage{xurl}
\externaldocument{supplement_short}

@@

We begin with an informal description of the allocation score and some examples illustrating its key characteristics in section \ref{sec:methods.overview}. In section \ref{sec:methods.detailed} we develop the allocation score more carefully, using a basic decision theoretic procedure for deriving proper scoring rules. (See section \ref{sec:proper} of the supplement for a definition of a proper scoring rule
and a more technical discussion of the procedure). In section \ref{sec:methods.related}, we note that another group of common scores including the quantile score, WIS, and CRPS, can also be derived from decision theoretic foundations \textemdash starting from a different decision making context.

@@

Suppose that a decision maker is tasked with determining how to allocate $K$ available units of a resource across $N$ locations.
If the decision maker is provided with a multivariate forecast $F$ where each marginal forecast distribution $F_i$ predicts resource need in a particular location, one option is to choose the resource allocation that minimizes the expected total unmet need according to the forecast.
We will give a more precise mathematical statement in section \ref{sec:methods.detailed}, but informally, the total expected unmet need according to the forecast is{}
\begin{align}
\sum_{i=1}^N \mathbb{E}_{F_i}[\text{unmet need in location $i$}], \label{eqn:informal_objective}
\end{align}
where the unmet need in a particular location is the difference between resource need in that location and the number of resources that were allocated there.
This allocation problem has an intuitively appealing solution: allocate so that the probabilities of need exceeding allocation in various locations are as close to each other as possible.
This will lead to the allocations provided by $F$ being quantiles of the marginal distributions $F_i$ for some \emph{single} probability level $\tau$ that is shared in common for all locations.

@@

We can formalize the preceding discussion with the following three-step procedure for developing scoring rules for probabilistic forecasts:
\begin{enumerate}
\item Specify a \emph{loss function} $s(x, y)$ that measures the loss associated with taking action $x$ when outcome $y$ eventually occurs.
\item Given a probabilistic forecast $F$, determine the \emph{Bayes act} $x^F$ that minimizes the expected loss under the distribution $F$.
\item The \emph{scoring rule} for $F$ calculates the score as the loss incurred when the Bayes act was used: $S(F, y) = s(x^F, y)$.
\end{enumerate}
This is a general procedure that may be applied in settings where it is possible to specify a quantitative loss function. We call a scoring rule obtained from this procedure a \emph{Bayes scoring rule} and in section \ref{sec:proper} of the 
supplement we show that they are proper by construction.

@@

To make this a more easily interpreted measure of forecast performance, we will adjust the raw score by subtracting the minimum loss
achievable by an \emph{oracle} allocator which has precise foreknowledge of the outcomes $y_i$.
When the oracle has sufficient resources to meet the total need, i.e., when $\sum_{i=1}^{N}y_i \leq K$, the oracle's loss is zero 
and allocation score coincides with the raw score.
On the other hand, when the oracle cannot cover all need and incurs a loss of $L \cdot \left(\sum_{i=1}^{N}y_i - K \right) > 0$, we adjust
the raw score by this loss.
The oracle-adjusted score can therefore be written as
\begin{align}
S_A(F, y; K)  &= S_A^{\text{raw}}(F, y; K) - L \cdot \max\left(0,\sum_{i=1}^{N}y_i - K\right) \\
&= L\left\{\sum_{i=1}^N \max(0, y_i - x_i^{F,K}) -  \max\left(0,\sum_{i=1}^{N}y_i - K\right)\right\}
\end{align}
The oracle adjustment aligns with a common theme in economic decision theory that \emph{opportunity loss} (often known as \emph{regret} or (negative) \emph{relative utility}) is often a more important quantity than absolute loss (see e.g., \cite{DIECIDUE201788}).

@@

In epidemiological forecasting, well-known proper scoring rules such as the log score or variations on the continuous ranked probability score (such as the weighted interval score, or WIS) have been frequently utilized to evaluate probabilistic forecasts.
Often models are ranked with respect to accuracy according to a particular score, but without reference to any underlying decision-making process for which that score was designed or from which it might be derivable as a Bayes scoring rule.
With careful thought and collaboration between modelers and public health officials, we argue that scores that are more aligned with public health decisions could be developed to inform specific problems.
We have demonstrated that forecast evaluation methods that are tied to a specific decision making context can yield model rankings that differ substantively from those based on standard measures of forecast skill.

@@

In practice, there are many potential users of forecasts with many different decision making problems.
Not all can be easily modeled as an effort to minimize the expectation of a transparent loss function.
And even those that are easily framed as expected loss minimization may differ enough that no single score would be 
appropriate for all users.
Ideally, targeted forecasting tools could be developed through close collaboration between modelers and public health officials.
However, this may only be possible in settings with sufficient staffing on both an analytics and a public health team.
Increasingly, collaborative modeling hubs are being used to generate ``one-size-fits-all'' forecasts for many locations at once.
In these settings, where tailored models are not available, it still could be possible to evaluate contributed models using a set of multiple scores to support public health end users in understanding the value of forecasts as an input to their particular decision making contexts.
% NGR: leaving this out for now as it feels like too much detail?
% This may be tricky to operationalize in the setting of a general forecast hub. It matters how you elicit and represent probabilistic forecasts (quantiles? samples? cdfs?).

@@

There are several important limitations to the current work.
The allocation score we developed here does not directly account for important considerations such as fairness or equity of allocations.
Nor does the proposed framework attempt to capture the broader context of decision making. For example, in practice it may be possible to increase the resource constraint $K$ by shifting funding from other disease mitigation measures.
We also note that in some settings, a ``successful'' epidemiological forecast may lead to policy decisions that change the distribution of the predicted outcome $Y$. Our framework would need considerable enhancements before being applicable to forecast evaluation beyond horizons
for which causal feedback can be neglected.

@@

In conclusion, we argue that the way modelers and policymakers view and evaluate forecasts should change depending on the specific decision-making context.
Defaulting to standard forecast evaluation metrics can mask the utility (or disutility) of certain forecasts, or lead to forecasts being used in decision making contexts very different from those for which they were intended.
New collaborative work between public health officials and modeling teams is needed to assess the value and relevance of the initial findings presented here, including real-time pilot studies or simulation exercises that could be used to inform further development of new or alternative scoring metrics.
We see this work as an initial overture for what we hope will grow to be a large, collaborative body of work more closely coupling applied epidemiological forecasting with public health decision making.

### 2023-12-20 alloscore_manuscript/alloscore-application-shorter.tex
Assorted main text (shorter) edits

\paragraph{Example 1} Suppose we have a forecast $F$ for need in two locations with $F_1 = \mathrm{Exp}(1 / \sigma_1)$ and $F_2 = \mathrm{Exp}(1 / \sigma_2)$, where $\sigma_1 = 1$ and $\sigma_2 = 4$. When the marginal forecasts are exponential distributions, it can be shown that the optimal allocation divides the available resources among the locations proportionally to the scale parameters $\sigma_i$ (see section
\ref{sec:bayes-quantiles} of the supplemental materials). If $K = 5$ units of the resource are available, the optimal allocation according to $F$ would be 1 unit of resources in location 1 and 4 units of resources in location 2. If, on the other hand, $K = 10$ units are available, we will allocate 2 units of resources to location 1 and 8 units to location 2. Figure~\ref{fig:exp_alloc_example} illustrates the situation.

@@

% latex table generated in R 4.3.1 by xtable 1.8-4 package
% Tue Dec 19 17:26:53 2023
\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \hline
model & AS & MWIS & IAS centered at 15k & IAS uniform \\ 
  \hline
CU-select & 669 & 133 & 774 & 326 \\ 
  COVIDhub-ensemble & 873 & 159 & 1067 & 438 \\ 
  USC-SI\_kJalpha & 995 & 91 & 1216 & 1097 \\ 
  JHUAPL-Gecko & 1034 & 164 & 1141 & 418 \\ 
  MUNI-ARIMA & 1084 & 169 & 1248 & 440 \\ 
  COVIDhub-trained\_ensemble & 1089 & 169 & 1271 & 823 \\ 
  COVIDhub-baseline & 1175 & 170 & 1317 & 535 \\ 
  JHUAPL-Bucky & 1358 & 102 & 1566 & 1214 \\ 
  JHUAPL-SLPHospEns & 1540 & 129 & 1604 & 1102 \\ 
  UVA-Ensemble & 2469 & 213 & 2635 & 2494 \\ 
   \hline
\end{tabular}
\caption{Comparison of allocation scores (AS), mean weighted interval scores (MWIS), and two varieties of Integrated Allocation Scores (IAS). All metrics are shown for 10 models that made forecasts of hospital admissions for 2022-01-03. Results are sorted by AS. For all metrics, lower scores indicate better accuracy.} 
\label{tab:multi-k-scores}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/thermometer-plot-1} \caption[Probabilistic forecasts for new hospital admissions and the inferred resource allocations for COVID-19 on January 3, 2022 for the states with the nine highest hospitalization counts]{Probabilistic forecasts for new hospital admissions and the inferred resource allocations for COVID-19 on January 3, 2022 for the states with the nine highest hospitalization counts. For each state, the dark black line shows the data observed when the forecast was made and the grey line shows eventually observed counts. The side-by-side shaded regions show the median (solid horizontal line) and 50\%, 80\% and 95\% prediction intervals for the two selected models. The forecasts were made for new hospitalizations on January 3, 2022 (vertical dashed line, with number of hospitalizations indicated by red horizontal line segment at the intersection of the dashed line and the grey line of data). The vertical bars with red and yellow shading show the allocations: the red bar goes from zero to the level of the observed number of hospitalizations while the yellow bar shows how many resources were suggested by the model to be allocated to that location. The allocations (yellow bar) may be more or less than the observed number, therefore exposed red bar indicates unmet need, while the yellow bar extending above the observed number of hospitalizations means that excess resources were suggested for that given location.}\label{fig:thermometer-plot}
\end{figure}

>>>>

\paragraph{Example 1} Suppose we have a forecast $F$ for need in two locations with $F_1 = \mathrm{Exp}(1 / \sigma_1)$ and $F_2 = \mathrm{Exp}(1 / \sigma_2)$, where $\sigma_1 = 1$ and $\sigma_2 = 4$. The means of these distributions are given by the scale parameters $\sigma_i$. When the marginal forecasts are exponential distributions, it can be shown that the optimal allocation divides the available resources among the locations proportionally to the scale parameters $\sigma_i$ (see section
\ref{sec:bayes-quantiles} of the supplemental materials). If $K = 5$ units of the resource are available, the optimal allocation according to $F$ would be 1 unit of resources in location 1 and 4 units of resources in location 2. If, on the other hand, $K = 10$ units are available, we will allocate 2 units of resources to location 1 and 8 units to location 2. Figure~\ref{fig:exp_alloc_example} illustrates the situation.

@@

% latex table generated in R 4.3.1 by xtable 1.8-4 package
% Wed Dec 20 04:09:40 2023
\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \hline
Model & AS & MWIS & IAS centered at 15k & IAS uniform \\ 
  \hline
CU-select & 669 & 133 & 774 & 326 \\ 
  COVIDhub-ensemble & 873 & 159 & 1067 & 438 \\ 
  USC-SI\_kJalpha & 995 & 91 & 1216 & 1097 \\ 
  JHUAPL-Gecko & 1034 & 164 & 1141 & 418 \\ 
  MUNI-ARIMA & 1084 & 169 & 1248 & 440 \\ 
  COVIDhub-trained\_ensemble & 1089 & 169 & 1271 & 823 \\ 
  COVIDhub-baseline & 1175 & 170 & 1317 & 535 \\ 
  JHUAPL-Bucky & 1358 & 102 & 1566 & 1214 \\ 
  JHUAPL-SLPHospEns & 1540 & 129 & 1604 & 1102 \\ 
  UVA-Ensemble & 2469 & 213 & 2635 & 2494 \\ 
   \hline
\end{tabular}
\caption{Comparison of allocation scores (AS), mean weighted interval scores (MWIS), and two varieties of Integrated Allocation Scores (IAS). All metrics are shown for 10 models that made forecasts of hospital admissions for 2022-01-03. Results are sorted by AS. For all metrics, lower scores indicate better accuracy.} 
\label{tab:multi-k-scores}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/thermometer-plot-1} \caption[Probabilistic forecasts for new hospital admissions and the inferred resource allocations for COVID-19 on January 3, 2022 for the states with the ten highest hospitalization counts]{Probabilistic forecasts for new hospital admissions and the inferred resource allocations for COVID-19 on January 3, 2022 for the states with the ten highest hospitalization counts. For each state, the dark black line shows the data observed when the forecast was made and the grey line shows eventually observed counts. The side-by-side shaded regions show the median (solid horizontal line) and 50\%, 80\% and 95\% prediction intervals for the two selected models. The forecasts were made for new hospitalizations on January 3, 2022 (vertical dashed line, with number of hospitalizations indicated by red horizontal line segment at the intersection of the dashed line and the grey line of data). The vertical bars with red and yellow shading show the allocations: the red bar goes from zero to the level of the observed number of hospitalizations while the yellow bar shows how many resources were suggested by the model to be allocated to that location. The allocations (yellow bar) may be more or less than the observed number, therefore exposed red bar indicates unmet need, while the yellow bar extending above the observed number of hospitalizations means that excess resources were suggested for that given location.}\label{fig:thermometer-plot}
\end{figure}
